Formalizing the Kolmogorov extension Theorem in Lean

### 2023-09-17 Manuscript/proof_outline.tex
accept most of the blue, rewrite one or two small passages

\begin{abstract}
  We present a formalization of the Kolmogorov extension theorem in
  Lean. This result is a main building block in the construction of
  the (distribution of) stochastic processes with arbitrary index
  sets. Our approach is based on mathlib, the mathematical library for
  Lean. On our way, we provide a formalization of the Caratheodory
  extension theorem, which allows to extend a set function on a
  semi-ring to a proper measure on the $\sigma$-algebra generated by
  the semi-ring. A slight generalization of the classical Kolmogorov
  extension theorem allows us to construct the (distribution of) a
  stochastic process on a complete and separable pseudo-metric space
  rather than a metric space. {\color{blue}As an example, we define a
    product probability measure with arbitrary index set.}
\end{abstract}

@@

\noindent
{\em Theorem of Ionescu-Tulcea and related results:}
While Kolmogorov's extension Theorem gives a result for arbitrary $\iota$, but requires some properties of the family $(\alpha_t)_{t\in\iota}$, the Theorem of Ionescu-Tulcea \cite[Theorem 8.24]{Kallenberg2020} can only deal with countable $\iota$, but has no restrictions on $(\alpha_t)_{t\in\iota}$. The proof of the latter, however, uses transition kernels (and an inductive construction of a content, which can be extended to a measure) rather than projective families.
However, the structure of the argument is similar and could use some of our results, like the proof of $\sigma$-additivity from continuity at the empty set (see Lemma~\ref{P:stetigmass}).
A formalization would complement our work. In addition, using the Theorem of Ionescu-Tulcea one can define infinitely many independent random variables (arbitrary $\iota$) on a joint probability space \cite[Theorem 8.24]{Kallenberg2020} {\color{blue} with fewer requirements on $(\alpha_t)_{t\in\iota}$ than in the example given in Section~\ref{ss:concl}}.
All of these results can be used in order to construct countably many independent random variables, which are frequently used in concrete constructions in probability theory. Examples are (one direction of) the Borel-Cantelli-Lemma \cite[Theorem 4.28]{Kallenberg2020}, Kolmogorov's 0-1-law \cite[Theorem 4.13]{Kallenberg2020}, random walks \cite[Chapter 12]{Kallenberg2020}, branching processes  \cite[Chapter 13]{Kallenberg2020}, and percolation  \cite[Chapter 2.4]{Klenke2013}, to name just a few. While the first two are already formalized in \leanline{mathlib}, it is only clear due to the Kolmorov extension theorem (or the Theorem of Ionesu-Tulcea) that infinitely many independent random variables can indeed be defined on a joint probability space. In this sense, our contribution is also important for giving more sense to already formalised parts of \leanline{mathlib}.

@@

In contrast to statements already part of \leanline{mathlib}, in the
sequel we \leanlinecolor{highlight} definitions and statements which
are part of our own code. In the formalization of the above, we use
\leanline{variable {ι : Type _} {α: ι → Type _}}, which fixes the
index set $\iota$ and all spaces $\alpha_t, t\in\iota$ as global
variables.  In Lean, the product space is then \leanline{∀ j : ι, α j}
and the product over a finite set \leanline{J} is \leanline{∀ j : J, α
  j}. {\color{blue} Note that \leanline{J} is a subtype of
  \leanline{ι} in this formulation.}

@@

\noindent
It is worth understanding the precise connection of
\leanlinecolor{isProjective} and
\leanlinecolor{isProjectiveMeasureFamily}. In the latter, the first
variable of \leanlinecolor{isProjective} is the family \leanline{P} of
finite measures for all finite subsets of \leanline{ι}. The second
variable is the functions which maps two sets \leanline{I J : Finset
  ι} and a proof \leanline{hJI} of \leanline{J ⊆ I} together with
\leanline{P I} (which is \leanline{μ} in the statement) to the image
measure on \leanline{J}, which is \leanline{μ.map (fun x : ∀ i : I, α
  i => fun j => x ⟨j, hJI j.2⟩)}.  The map defined by the
\leanline{=>}-notation maps every $(x_i)_{i\in I}$ to a function of
\leanline{j}, whose type is the subtype {\color{blue}\leanline{j} of}
\leanline{I}, consisting of a value and a proof of \leanline{J ⊆
  I}. In other words, this is $(x_j)_{j\in J}$.

@@

\noindent
\subsection{Carathéodory's extension theorem}
Measures are set functions defined on a $\sigma$-algebra, satisfying
some properties which we recall below. {\color{blue}Frequently,
  constructively} defining such a measure on the full $\sigma$-algebra
is not possible, but defining a set function on a semi-ring is
possible. As an example, consider Lebesgue-measure on $\mathbb R$,
with the Borel $\sigma$-algebra on $\mathbb R$, which is
$\sigma(\mathcal O)$, where $\mathcal O$ is the set of open sets, and
$\sigma(\mathcal O)$ is the smallest $\sigma$-algebra containing all
open sets. Since this set system is defined only abstractly, it is
hard to know which volume each of these sets should be assigned to at
first sight. However, the volume of an interval is easy, since it may
be defined by the length of the interval. So, in order to construct
measures from a set function $m$ on a (semi-)ring $\mathcal H$
(e.g.\ the set of all semi-open intervals), it has been a fundamental
insight of Carathéodory that one may start by defining a set function
(outer measure) $\mu$ on $2^\alpha$, and then show $\mu$ extends $m$
and defines a measure on {\color{blue}the $\sigma$-algebra of
  measurable sets (see Theorem~\ref{T:cara} below), which contains}
$\sigma(\mathcal H)$.
%(Note that the set of semi-open intervals
%generates the Borel $\sigma$-algebra on $\mathbb R$.)
We will follow this abstract construction, and start by stating some
basic concepts.

@@

\noindent
{\color{blue} Such compact systems are important since they allow a
  proof of $\sigma$-additivity of a content on a ring, which is the
  missing piece for applying the Carathéodory Theorem in the proof of
  the Kolmogorov extension theorem.}

@@

Note that the assumption \leanline{hP_inner} from above translates
directly to inner regularity of the \leanlinecolor{kolContent}, which
is defined through the projective family \leanline{P}. Moreover, since
the Carathéodory extension theorem requires $\sigma$-subadditivity
(rather than $\sigma$-additivity), we can use
\leanlinecolor{sigma_subadditive_of_sigma_additive} {\color{blue}(see
  Lemma~\ref{P:stetigmass}, 1.$\Rightarrow$2.)} in order to show that
the assumptions of the last result also imply $\sigma$-subadditivity;
this gives
\leanlinecolor{kolContent_sigma_subadditive_of_innerRegular}.

@@

In the next lemma, we will need that a closed subset of a compact set
is compact. (See \leanline{isCompact_of_isClosed_subset}.) We describe
the formalization of the next lemma only in conjunction with
Lemma~\ref{L:relcoPol}. {\color{blue} We omit the proof, since the
  only interesting part is an application of inner regularity from
  Lemma~\ref{l:pseude1}.}

@@

\begin{proof}
 Since $\alpha$ is separable, there is a subset $\{x_1, x_2,...\}
 \subseteq_c \alpha$ with closure $\alpha$. By separability, for each
 $n\in \mathbb N$, $\alpha = \bigcup_{k=1}^\infty
 B_{\varepsilon_n}(x_k)$. Since $\mu$ is continuous from above in
 $\emptyset$ (see Lemma~\ref{P:stetigmass})
  \[
  0 = \mu\Big(\Big(\bigcup_{k=1}^\infty B_{\varepsilon_n}(x_k)\Big)^c
  \Big) = \lim_{N\to\infty} \mu\Big( \Big(\bigcup_{k=1}^N
  B_{\varepsilon_n}(x_k)\Big)^c\Big).
  \]
  Let $\delta_n\downarrow 0$ be summable with $\sum_{n=1}^\infty
  \delta_n = 1$ (e.g.\ $\delta_n = 2^{-n}$). Then, there is some
  $N_n\in\mathbb N$ with $\mu\Big( E\setminus \bigcup_{k=1}^{N_n}
  B_{\varepsilon_n}(x_k)\Big) < \varepsilon \delta_n$. Now take
  \[
    A := \bigcap_{n=1}^\infty \bigcup_{k=1}^{N_n}
    B_{\varepsilon_n}(x_k),
  \]
  which by construction is totally bounded (for $r>0$ take $n$ such
  that $\varepsilon_n < r$, such that $B_r(x_1),...,B_r(x_{N_n})$
  cover $A$), and therefore has a compact closure as noted
  in the lines directly preceding the lemma. In
  addition, by $\sigma$-sub-additivity of $\mu$,
  \begin{align*}
     \mu((\overline A)^c) & \leq \mu(A^c) = \mu\Big(
     \bigcup_{n=1}^\infty \Big(\Big(\bigcup_{k=1}^{N_n}
     B_{\varepsilon_n}(x_k)\Big)^c\Big)\Big) \\ & \leq
     \sum_{n=1}^\infty \mu\Big( \Big(\bigcup_{k=1}^{N_n}
     B_{\varepsilon_n}(x_k)\Big)^c\Big)< \varepsilon,
  \end{align*}
 and we are done.
\end{proof}

Let us combine the last two results in one formalized result.  This
the gives the desired statement that a measure on a second countable,
complete extended pseudo-metric space is inner regular with respect to
closed compact sets.

@@

\section*{Acknowledgements}
{\color{blue}PP ist partly funded by the Vector foundation, grant {\em
    LeanAI (P2022-0241)}}.

>>>>

\begin{abstract}
  We present a formalization of the Kolmogorov extension theorem in
  Lean. This result is a main building block in the construction of
  the (distribution of) stochastic processes with arbitrary index
  sets. Our approach is based on mathlib, the mathematical library for
  Lean. On our way, we provide a formalization of the Caratheodory
  extension theorem, which allows to extend a set function on a
  semi-ring to a proper measure on the $\sigma$-algebra generated by
  the semi-ring. A slight generalization of the classical Kolmogorov
  extension theorem allows us to construct the (distribution of) a
  stochastic process on a complete and separable pseudo-metric space
  rather than a metric space. As an example, we define a
  product probability measure with arbitrary index set.
\end{abstract}

@@

\noindent
{\em Theorem of Ionescu-Tulcea and related results:}
While Kolmogorov's extension Theorem gives a result for arbitrary $\iota$, but requires some properties of the family $(\alpha_t)_{t\in\iota}$, the Theorem of Ionescu-Tulcea \cite[Theorem 8.24]{Kallenberg2020} can only deal with countable $\iota$, but has no restrictions on $(\alpha_t)_{t\in\iota}$. The proof of the latter, however, uses transition kernels (and an inductive construction of a content, which can be extended to a measure) rather than projective families.
However, the structure of the argument is similar and could use some of our results, like the proof of $\sigma$-additivity from continuity at the empty set (see Lemma~\ref{P:stetigmass}).
A formalization would complement our work. In addition, using the Theorem of Ionescu-Tulcea one can define infinitely many independent random variables (arbitrary $\iota$) on a joint probability space \cite[Theorem 8.24]{Kallenberg2020} with fewer requirements on $(\alpha_t)_{t\in\iota}$ than in the example given in Section~\ref{ss:concl}.
All of these results can be used in order to construct countably many independent random variables, which are frequently used in concrete constructions in probability theory. Examples are (one direction of) the Borel-Cantelli-Lemma \cite[Theorem 4.28]{Kallenberg2020}, Kolmogorov's 0-1-law \cite[Theorem 4.13]{Kallenberg2020}, random walks \cite[Chapter 12]{Kallenberg2020}, branching processes  \cite[Chapter 13]{Kallenberg2020}, and percolation  \cite[Chapter 2.4]{Klenke2013}, to name just a few. While the first two are already formalized in \leanline{mathlib}, it is only clear due to the Kolmorov extension theorem (or the Theorem of Ionesu-Tulcea) that infinitely many independent random variables can indeed be defined on a joint probability space. In this sense, our contribution is also important for giving more sense to already formalised parts of \leanline{mathlib}.

@@

In contrast to statements already part of \leanline{mathlib}, in the
sequel we \leanlinecolor{highlight} definitions and statements which
are part of our own code. In the formalization of the above, we use
\leanline{variable {ι : Type _} {α: ι → Type _}}, which fixes the
index set $\iota$ and all spaces $\alpha_t, t\in\iota$ as global
variables.  In Lean, the product space is then \leanline{∀ j : ι, α j}
and the product over a finite subset \leanline{J} of \leanline{ι} is \leanline{∀ j : J, α
  j}.

@@

\noindent
It is worth understanding the precise connection of
\leanlinecolor{isProjective} and
\leanlinecolor{isProjectiveMeasureFamily}. In the latter, the first
variable of \leanlinecolor{isProjective} is the family \leanline{P} of
finite measures for all finite subsets of \leanline{ι}. The second
variable is the functions which maps two sets \leanline{I J : Finset
  ι} and a proof \leanline{hJI} of \leanline{J ⊆ I} together with
\leanline{P I} (which is \leanline{μ} in the statement) to the image
measure on \leanline{J}, which is \leanline{μ.map (fun x : ∀ i : I, α
  i => fun j => x ⟨j, hJI j.2⟩)}.  The function used in that image
measure maps every $(x_i)_{i\in I}$ to $(x_i)_{i\in J}$.
%map defined by the
%\leanline{=>}-notation maps every $(x_i)_{i\in I}$ to a function of
%\leanline{j}, whose type is the subtype {\color{blue}\leanline{j} of}
%\leanline{I}, consisting of a value and a proof of \leanline{J ⊆
%  I}. In other words, this is $(x_j)_{j\in J}$.

@@

\noindent
\subsection{Carathéodory's extension theorem}
Measures are set functions defined on a $\sigma$-algebra, satisfying
some properties which we recall below. Frequently,
constructively defining such a measure on the full $\sigma$-algebra
is not possible, but defining a set function on a semi-ring is
possible. As an example, consider Lebesgue-measure on $\mathbb R$,
with the Borel $\sigma$-algebra on $\mathbb R$, which is
$\sigma(\mathcal O)$, where $\mathcal O$ is the set of open sets, and
$\sigma(\mathcal O)$ is the smallest $\sigma$-algebra containing all
open sets. Since this set system is defined only abstractly, it is
hard to know which volume each of these sets should be assigned to at
first sight. However, the volume of an interval is easy, since it may
be defined by the length of the interval. So, in order to construct
measures from a set function $m$ on a (semi-)ring $\mathcal H$
(e.g.\ the set of all semi-open intervals), it has been a fundamental
insight of Carathéodory that one may start by defining a set function
(outer measure) $\mu$ on $2^\alpha$, and then show $\mu$ extends $m$
and defines a measure on the $\sigma$-algebra of
  measurable sets (see Theorem~\ref{T:cara} below), which contains
$\sigma(\mathcal H)$.
%(Note that the set of semi-open intervals
%generates the Borel $\sigma$-algebra on $\mathbb R$.)
We will follow this abstract construction, and start by stating some
basic concepts.

@@

\noindent
Such compact systems are important since they allow a
proof of $\sigma$-additivity of a content on a ring, which is the
missing piece for applying the Carathéodory Theorem in the proof of
the Kolmogorov extension theorem.

@@

Note that the assumption \leanline{hP_inner} from above translates
directly to inner regularity of the \leanlinecolor{kolContent}, which
is defined through the projective family \leanline{P}. Moreover, since
the Carathéodory extension theorem requires $\sigma$-subadditivity
(rather than $\sigma$-additivity), we can use
\leanlinecolor{sigma_subadditive_of_sigma_additive} (see
  Lemma~\ref{P:stetigmass}, 1.$\Rightarrow$2.) in order to show that
the assumptions of the last result also imply $\sigma$-subadditivity;
this gives
\leanlinecolor{kolContent_sigma_subadditive_of_innerRegular}.

@@

In the next lemma, we will need that a closed subset of a compact set
is compact. (See \leanline{isCompact_of_isClosed_subset}.) We describe
the formalization of the next lemma only in conjunction with
Lemma~\ref{L:relcoPol}. We omit the proof, since the
  only interesting part is an application of inner regularity from
  Lemma~\ref{l:pseude1}.

@@

\begin{proof}
 Since $\alpha$ is separable, there is a subset $\{x_1, x_2,...\}
 \subseteq_c \alpha$ with closure $\alpha$. By separability, for each
 $n\in \mathbb N$, $\alpha = \bigcup_{k=1}^\infty
 B_{\varepsilon_n}(x_k)$. Since $\mu$ is continuous from above in
 $\emptyset$ (see Lemma~\ref{P:stetigmass})
  \[
  0 = \mu\Big(\Big(\bigcup_{k=1}^\infty B_{\varepsilon_n}(x_k)\Big)^c
  \Big) = \lim_{N\to\infty} \mu\Big( \Big(\bigcup_{k=1}^N
  B_{\varepsilon_n}(x_k)\Big)^c\Big).
  \]
  Let $\delta_n\downarrow 0$ be summable with $\sum_{n=1}^\infty
  \delta_n = 1$ (e.g.\ $\delta_n = 2^{-n}$). Then, there is some
  $N_n\in\mathbb N$ with $\mu\Big( E\setminus \bigcup_{k=1}^{N_n}
  B_{\varepsilon_n}(x_k)\Big) < \varepsilon \delta_n$. Now take
  \[
    A := \bigcap_{n=1}^\infty \bigcup_{k=1}^{N_n}
    B_{\varepsilon_n}(x_k),
  \]
  which by construction is totally bounded (for $r>0$ take $n$ such
  that $\varepsilon_n < r$, such that $B_r(x_1),...,B_r(x_{N_n})$
  cover $A$), and therefore has a compact closure as noted
  in the lines directly preceding the lemma. Finally, by $\sigma$-sub-additivity of $\mu$,
  \begin{align*}
     \mu((\overline A)^c) & \leq \mu(A^c) = \mu\Big(
     \bigcup_{n=1}^\infty \Big(\Big(\bigcup_{k=1}^{N_n}
     B_{\varepsilon_n}(x_k)\Big)^c\Big)\Big) \\ & \leq
     \sum_{n=1}^\infty \mu\Big( \Big(\bigcup_{k=1}^{N_n}
     B_{\varepsilon_n}(x_k)\Big)^c\Big)< \varepsilon
     \: .
  \end{align*}
\end{proof}

Combining the last two results, we obtain the desired statement that a measure on a second countable,
complete extended pseudo-metric space is inner regular with respect to
closed compact sets.

@@

\section*{Acknowledgements}
PP ist partly funded by the Vector foundation, grant
{\em LeanAI (P2022-0241)}.
RD acknowledges the support of the French Ministry of Higher Education and Research and the Hauts-de-France region.
